---
title: "Homework 5"
author: "Skylar Liu"
date: "2024-02-04"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_template$set(figure1 = list(fig.height = 3.7, fig.width = 5),
                         figure2 = list(fig.height = 3.7, fig.width = 8))
```

Homework #5, Stat 660, Spring 2024, Due Class #7, February 7, 2024

1. Do a similar analysis about heteroscedasticity for each choice of k as was done in Lecture 5. 
Plot the fitted absolute residuals against either the predicted values or against X, your
choice. Put them on one graph, and display.

```{r absolute residuals, opts.label = "figure1", fig.align="center"}
# clear workspace
rm(list = ls())
# set the seed
set.seed(382957)
library(HRW)
library(mgcv)

# import data
fossil = read.csv("~/660 - Flexible Regression/Homework/Homework3/fossil.csv")
X = fossil$Age
Y = fossil$Strontium.Ratio

# Fit a default gam to the data
gam_default = gam(Y~s(X,bs="cr"))

# mgcv fit with K = 4 knots
gam4 = gam(Y~s(X,bs="cr",k=4))
# mgcv fit with K = 8 knots
gam8 = gam(Y~s(X,bs="cr",k=8))
# mgcv fit with K = 23 knots
gam23 = gam(Y~s(X,bs="cr",k=23))

# absolute residual plot against predicted values
ord = sort(X, index.return = T)$ix
plot(X, abs(gam_default$residuals),
     ylab="Absolute Residuals", 
     main="Absolute Residuals of Fossil data for K=4,8,23",
     xlab="Age (X)")
y4    = abs(gam4$residuals)
y8   = abs(gam8$residuals)
y23    = abs(gam23$residuals)
gam4_abs = gam(y4~s(X,bs="cr"))
gam8_abs = gam(y8~s(X,bs="cr"))
gam23_abs = gam(y23~s(X,bs="cr"))
lines(X[ord],fitted(gam4_abs)[ord],lwd=3,col="blue")
lines(X[ord],fitted(gam8_abs)[ord],lwd=3,col="red")
lines(X[ord],fitted(gam23_abs)[ord],lwd=3,col="green")
legend("topright", legend = c("K=4", "K=8", "K=23"), col=c("blue", "red", "green"),lty=1, cex=0.8)
```
2. Display the ratio of the maximum fitted absolute residual to the minimum fitted absolute
residual, for each case.

```{r abs ratios}
# ratio of the maximum fitted absolute residual to the minimum fitted absolute residual
ratio_k4 = max(abs(fitted(gam4_abs))) / min(abs(fitted(gam4_abs)))
ratio_k8 = max(abs(fitted(gam8_abs))) / min(abs(fitted(gam8_abs)))
ratio_k23 = max(abs(fitted(gam23_abs))) / min(abs(fitted(gam23_abs)))
cat('k = 4 fit, maximum to minimum ratio = ',ratio_k4,"\n")
cat('k = 8 fit, maximum to minimum ratio = ',ratio_k8,"\n")
cat('k = 23 fit, maximum to minimum ratio = ',ratio_k23,"\n")
```
> All of these fits fail the rule of thumb, since the ratios are all above 3. K=8 and K=23 are just over the line while K=4 is well over 3.

4. If any choice of k fails my rule of thumb, describe in words what that means. Remember,
it is possible that the heteroscedasticity can occur in the middle of the X values, and not
merely at the end.

> Since all of these values fail the rule of thumb, there is heteroscedasticity somewhere in the model and we cannot trust the confidnce intervals.

5. Give a verbal description of what k is.

> K is the number of knots, which are points where the slope appears to change on a continuous fitted line. Increased knots increase flexibility and wiggles in the line.

6. Load the trees data set: install.packages("trees", dependencies=TRUE). Then
library(trees). Ignore any messages. The response Y is volume of the tree, and the
predictors are girth and height of the tree. Run mgcv::gam with volume as the response,
height entering the model linearly, and girth entering as a cubic Bspline with k=5 (there
are only 32 or so data points, so too many basis functions are silly). Show the summary
output and plot the model object (just guess).

```{r trees, opts.label = "figure1", fig.align="center"}
# import the tree package
library(tree)

# define variables
volume = trees$Volume
height = trees$Height
girth  = trees$Girth

# fit the tree data
gam_spline = gam(volume ~ height + s(girth,bs="cr", k=5))
summary(gam_spline)

# plot the model
plot(gam_spline, 
     main="Tree Spline Model",
     xlab="Volume", ylab="fitted(gam_spline)", cex=2)
```

7. Rerun the model, except this time making girth a linear predictor. Compare the two models
using an anova statement. Try anova(‘Linear Fit Name”,”Spline fit Name”,type=”?”), and
display the p-value. With such a small sample size, do not be surprised if the p-value is not
< 0.05.

```{r mgcv fit}
# fit using girth as a linear predictor
gam_linear = gam(volume ~ height + girth)

# anova comparison
anova(gam_linear, gam_spline, test = "Chisq")
cat('p-value = ',anova(gam_linear, gam_spline, test = "Chisq")$"Pr(>Chi)"[2],"\n")
```
