---
title: "Homework 4"
author: "Skylar Liu"
date: "2024-02-03"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_template$set(figure1 = list(fig.height = 3.7, fig.width = 5),
                         figure2 = list(fig.height = 3.7, fig.width = 8))
```

Homework #4, Stat 660, Spring 2024, Class #6, Monday February 5, 2024

1. Display and study the scatterplot of these data. What features of the data look interesting
to you? Do this before answering the other questions.

```{r scatterplot, opts.label = "figure1", fig.align="center"}
# clear workspace
rm(list = ls())
# set the seed
set.seed(382957)

# import data
fossil = read.csv("~/660 - Flexible Regression/Homework/Homework3/fossil.csv")
X = fossil$Age
Y = fossil$Strontium.Ratio

# scatterplot X vs Y
plot(X, Y, type="p",pch='*',main="Fossil Age vs. Strontium Ratio",
     xlab="Age",ylab="Strontium Ratio",cex=2)
```
> The pattern of the data is interesting to me, particularly how the Strontium Ratio dips around 114 years of Age. It appears to be in a general cubic shape.

2. Fit the fossil data using the default version of smooth.spline.

>a. Get and save the model object. You use something like myspline =
smooth.spline(…). The model object is myspline.

```{r smooth spline}
# Get the smooth.spline function
library(HRW)

# fit the smooth.spline model
myspline = smooth.spline(X, Y)
```

>b. Add the fitted line to the scatter plot of the data and display the resulting plot.

```{r fitted line, opts.label = "figure1", fig.align="center"}
# use the predict function to setup the line
mypred = predict(myspline)

# plot the scatter plot and fitted line
plot(X, Y, type="p",pch='*',main="Fossil Data With Spline Fit",
     xlab="Age",ylab="Strontium Ratio",cex=2)
lines(mypred$x,mypred$y,lwd=3,col="red")
```

3. Run the mgcv fit to the data with the default number of knots (K=8) and with both K = 4
and K = 23 knots and using the cubic spline option as I have done. Save the model fit
objects, e.g., gam4, gam8 and gam23.

>a. Which fits are statistically significant? Be sure to quote the p-values for all three.

```{r mgcv fit}
# Get the mgcv package
library(mgcv)

# mgcv fit with K = 4 knots
gam4 = gam(Y~s(X,bs="cr",k=4))
summary(gam4)

# mgcv fit with K = 8 knots
gam8 = gam(Y~s(X,bs="cr",k=8))
summary(gam8)

# mgcv fit with K = 23 knots
gam23 = gam(Y~s(X,bs="cr",k=23))
summary(gam23)
```
> The p-value for all three knot values (k = 4, 8, and 23) is 2^-16. Since this value is less than alpha=.05, they are all statistically significant.

>b. Plot the fits with the data points on one graph ONLY and submit that graph.

```{r mgcv fit plot, opts.label = "figure1", fig.align="center"}
# sort the data
ord = sort(X, index.return = T)$ix

# plot the scatter plot and fitted lines
plot(X, Y, type="p",pch='*',main="Fossil Data With GAM Fits",
     xlab="Age",ylab="Strontium Ratio",cex=2)
lines(X[ord], fitted(gam4)[ord], lwd=3, col="blue")
lines(X[ord], fitted(gam8)[ord], lwd=3, col="red")
lines(X[ord], fitted(gam23)[ord], lwd=3, col="green")
legend("bottomleft", legend = c("K=4", "K=8", "K=23"), col=c("blue", "red", "green"),lty=1, cex=0.8)
```
>c. Do the fits agree more or less with your answer to Question 1? Why or why not?

> In general, the plots create a cubic fit consistent with my observations. The K=23 fit deviates slightly, with bumps around age 100. 

4. What are the effective degrees of freedom for each mgcv fit?

```{r EDF}
# check the effective degrees of freedom for each mgcv fit
gam.check(gam4)
gam.check(gam8)
gam.check(gam23)
```

> EDF of gam4: 3
EDF of gam8: 6.5
EDF of gam23: 11.5

```{r lambdas}
# lambda of each mgcv fit
gam4$sp
gam8$sp
gam23$sp
```
> gam4: lambda = 0.01212764 
gam8: lambda = 0.5889037 
gam23: lambda = 7.493726 

6. Tell me whether or not the p-value for each choice of K is < 0.10. Cite those p-values. If
any are < 0.10, then explain intuitively from your graphs why that number of basis
functions is inadequate.

> K = 4: p-value = <2e-16 < 0.10
K = 8: p-value = 0.07 < 0.10
K = 23: p-value = 0.68 < 0.10

> K=4 and K=8 p-values are < 0.10, suggesting an inadequate number of knots. These number of basis functions fail to pick up the bump that is present in K=23, around age 100.

