---
title: "Homework 10"
author: "Skylar Liu"
date: "2024-03-17"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_template$set(figure1 = list(fig.height = 3.7, fig.width = 5),
                         figure2 = list(fig.height = 3.7, fig.width = 8))
```

Homework #10, Stat 660, Spring 2024, Due Tuesday March 19 by 11:59PM

1. We have talked about the spinal bone mineral density data, with the random intercept case. 
I stated that I think, or perhaps hope, that the id numbers should not have to be 1,2,3… I 
also think/hope that the idnumbers do not need to be ordered. 
It is generally the case that data sets are “cleaned” and there is a code book that converts 
the actual ids to 1,2,3,…,n. This helps de-identify data and helps preserve anonymity.

>a. Test this out in the spinal bone mineral density data, by defining a new variable, 
femSBMD$idnum2 = 2*femSBMD$idnum.

```{r new variable}
rm(list = ls())
set.seed(382957)
options(repos = list(CRAN="http://cran.rstudio.com/"))

# load libraries
library(HRW)
library(gamm4)

# import data
femSBMD = read.csv("~/660 - Flexible Regression/Homework/Homework10/femSBMD.csv")

# add new variable
femSBMD$idnum2 = 2*femSBMD$idnum
```

>b. Then rerun the gamm4::gamm4 given in class to see if you get the same results. I 
think/hope you will.

```{r gamm4}
# fit from class
fitclass <- gamm4(spnbmd ~ s(age,k=10,bs="cr") + black 
                + hispanic + white,
              random= ~(1|idnum),data = femSBMD)

fitLclass <- gamm4(spnbmd ~ I(age) + black + hispanic + white,
              random= ~(1|idnum),data = femSBMD)

# fit with new variable
fit <- gamm4(spnbmd ~ s(age,k=10,bs="cr") + black 
                + hispanic + white,
              random= ~(1|idnum2),data = femSBMD)

fitL <- gamm4(spnbmd ~ I(age) + black + hispanic + white,
              random= ~(1|idnum2),data = femSBMD)

# comparison
anova(fitclass$mer,fit$mer)
anova(fitLclass$mer,fitL$mer)
```
> The p-value from the anova for comparing both the fit and fitL from class compared to with the new variable is 0, showing there is no difference between these models. 

2. Get the data set pigWeights.csv from Canvas. The variable weight is the response, and the 
variable num.weeks is the date of the repeated measures.

>a. Display the lattice plot from library(lattice). Use the example from the spinal bone 
mineral density data to do this. I covered this in class, but in this case there is only 
1 population and no ANCOVA.

```{r pig data}
# import data 
pigWeights = read.csv("~/660 - Flexible Regression/Homework/Homework10/pigWeights.csv")

# load packages
library(lattice)

# lattice plot
pigWeightsvis <- xyplot(weight ~ num.weeks,
                     group = id.num,as.table = TRUE,
                     data = pigWeights,
                     strip = strip.custom(par.strip.text
                                          = list(cex = 1.5)),
                     par.settings = list(layout.heights
                                         =list(strip=1.6)),
                     scales = list(cex = 1.25),
                     xlab = list("weeks",cex = 1.5),
                     ylab = list(expression(paste(
                       "weight (kg)")),
                       cex = 1.5),
                     panel = function(x,y,subscripts,groups)
                     {  
                       panel.grid() 
                       panel.superpose(x,y,subscripts,groups,
                                       type = "b",pch = 16,lwd = 2)
                     })
plot(pigWeightsvis)
```

>b. Looking at the data, do you think a random intercept model holds for these data? 
Why or why not? You might want to look at Lecture 15 where I described the means 
and variances of a random intercept model. It is a subjective call, but just answer 
it.

> Each line seems to have the same shape and linear slope with minimal changes of subject to subject variance, so I believe a random intercept model holds for this data.

>c. Fit the random intercept model with num_weeks modeled as a spline. Do a summary 
and show your results. Show the between-person variance of the intercept and the 
within-person variance of the random errors. You may use either mgcv::gamm or 
gamm4::gamm4. They should be similar because gamm and gamm4 are 
theoretically justified in this family=gaussian case

```{r random intercept model}
# random intercept gamm
fit <- gamm4(weight ~ s(num.weeks,k=9,bs="cr"),
              random= ~(1|id.num),data = pigWeights)
summary(fit$mer)
```

> The between-person variance of the intercept is 15.152 while the within-person variance of the random errors is 4.298.

>d. Using anova() in gamm4::gamm4, to test whether a spline is needed as compared 
to a linear and a quadratic effect. 

```{r spline test}
# test whether a spline is needed
fitL <- gamm4(weight ~ I(num.weeks),
              random= ~(1|id.num),data = pigWeights)
fitQ <- gamm4(weight ~ I(num.weeks) + I(num.weeks^2),
              random= ~(1|id.num),data = pigWeights)

anova(fit$mer,fitL$mer)
anova(fit$mer,fitQ$mer)
```
> Compared to both linear and quadratic fits, a spline is necessary only for the quadratic fit as the p-value is very small (~0). A spline may not be necessary compared to the linear fit sincce the p-value is slightly above an alpha of 0.05 (p-value = 0.098)

e. Compare the quadratic and linear fits as well.

```{r fit comparison}
# comparing linear vs quadratic fit
anova(fitQ$mer,fitL$mer)
```
> The resulting p-value is 0.39 when comparing a quadratic fit to a linear fit, suggesting that the linear fit is the better fit.

>f. As in the spinal bone mineral density data, plot the fixed effects function against 
num.weeks, and include a pointwise 95% confidence interval for it.

```{r fixed effects}
# setup predictions
ng <- 432
num.weeks = seq(min(pigWeights$num.weeks) , max(pigWeights$num.weeks),length=ng)
pred <- predict(fit$gam,newdata=data.frame(num.weeks = num.weeks),se.fit=TRUE)
lowdirg <- pred$fit - qnorm(0.975) * pred$se.fit
uppdirg <- pred$fit + qnorm(0.975) * pred$se.fit
ymin = min(min(pred$fit))
ymax = max(max(pred$fit))

# Plot fixed effects function against num.weeks with 95% confidence intervals
plot(0, type = "n", 
     xlab = "Number of Weeks", 
     ylab = "Fitted values", 
     main = "Pig Weight Data",
xlim = c(min(num.weeks), max(num.weeks)), ylim = c(ymin, ymax))
polygon(c(num.weeks, rev(num.weeks)), c(lowdirg, rev(uppdirg)), col = "palegreen",
border = FALSE)
lines(num.weeks, pred$fit, col = "red", lwd = 2)
```

>g. Since you have already computed var(U) and var(epsilon), what is the estimated within-person correlation for this model?

```{r correlation}
# run the fit summary again
summary(fit$mer)

# within person correlation calculation
15.152336/(15.152336 + 4.297895)
```

> The within-person correlation for this model is 0.779.
