---
title: "Homework 9"
author: "Skylar Liu"
date: "2024-03-06"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_template$set(figure1 = list(fig.height = 3.7, fig.width = 5),
                         figure2 = list(fig.height = 3.7, fig.width = 8))
```

1. Install the package AER. Then get the library(AER), then data(HousePrices) and then
library(mgcv)

```{r initiate}
rm(list = ls())
set.seed(382957)
options(repos = list(CRAN="http://cran.rstudio.com/"))

#install and load data and packages
install.packages('AER')
library(AER)
data("HousePrices")
library(mgcv)
```

2. Run the code
>fitGaussAM = gam(price ~ s(lotsize,bs="cr",k=105)
+ bedrooms
+ factor(bathrooms)
+ factor(stories)
+ factor(driveway)
+ factor(recreation)
+ factor(fullbase)
+ factor(gasheat)
+ factor(aircon)
+ garage
+ factor(prefer),
data = HousePrices,family = gaussian)
summary(fitGaussAM)

Is lot size statistically significant at p=0.05?

```{r fitGaussAM}
# run example code
fitGaussAM = gam(price ~ s(lotsize,bs="cr",k=105)
                + bedrooms
                + factor(bathrooms)
                + factor(stories)
                + factor(driveway)
                + factor(recreation)
                + factor(fullbase)
                + factor(gasheat)
                + factor(aircon)
                + garage
                + factor(prefer),
                data = HousePrices,family = gaussian)
summary(fitGaussAM)
```
> The p-value of lot size is 2^-16, so it is significant at the alpha=0.5 level.

3. Use the anova function to test whether the spline model is better than just modeling lotsize
as a linear term.

```{r spline test}
# Lotsize as linear term
fitGaussFull = gam(price ~ lotsize
                + bedrooms
                + factor(bathrooms)
                + factor(stories)
                + factor(driveway)
                + factor(recreation)
                + factor(fullbase)
                + factor(gasheat)
                + factor(aircon)
                + garage
                + factor(prefer),
                data = HousePrices,family = gaussian)

# Test lotsize spline vs no spline
anova(fitGaussAM,fitGaussFull,test="Chisq")
```
> The p-value of the anova is 2.84^-5 which is statistically significatn at the 0.05 level, indicating that the spline model is better than no spline.

4. Use gam.check(fitGaussAM) to see if you have enough basis functions. Do you? 

```{r gam check}
# Check basis functions
gam.check(fitGaussAM)
```

> The p-value is large at 0.97, indicating that there are enough basis functions present.

5. Consider a house with a lot size of 5000 square feet, three bedrooms, two bathrooms, two
stories, a driveway, no recreation room, a finished basement, hot water heating, no air
conditioning, two garage places, and located outside of the preferred neighborhood of
Windsor, Canada. Predict the mean price of all houses under those constraints.

```{r prediction}
# Set up prediction parameters
example <- data.frame(lotsize = 5000,
                      bedrooms = 3,
                      bathrooms = 2,
                      stories = 2,
                      driveway = "yes",
                      recreation = "no",
                      fullbase = "yes",
                      gasheat = "yes",
                      aircon = "no",
                      garage = 2,
                      prefer = "no")

# Predict mean price of all houses with give constraints
predicted_price <- predict(fitGaussAM, newdata = example, type = "response", se.fit = TRUE)
mean = predicted_price$fit
se   = predicted_price$se.fit
mean
```
> The predicted mean housing price is $91,617.43.

6. Find an approximate 95% confidence interval for the mean price above.

```{r CI}
# 95% CI for the mean housing price
n = length(HousePrices$price)
t = qt(0.975, df = n-2)
upperCI = mean + t * se
lowerCI = mean - t * se
c(lowerCI, upperCI)
```
> The 95% CI for housing prices is ($81,561.41, $101,673.45)

7. Now restart R and and install the packages aplore3 and gam. Make sure you detach mgcv
before doing this.
a. Add in library(aplore3); library(gam); data(icu); help(icu).

```{r restart}
rm(list = ls())

detach("package:mgcv", unload=TRUE)
library(aplore3)
library(gam)

data(icu)
help(icu)
```

8. Let Y be the response that the patient dies. The help file will help you figure out what this
variable is. Run stepwise regression (step.Gam) with possible splines for age (age), heart
rate (hra) and systolic blood pressure (sys), with 3 EDF. The binary predictor (no spline)
is gender.

```{r stepwise}
# Make Y numeric
icu$sta = as.numeric(icu$sta == "Died")
Y = icu$sta

# Run a baseline linear regression
baseGam = gam:::gam(Y ~ as.factor(gender) 
                    + age 
                    + hra 
                    + sys, 
                    family = binomial,
                    data=icu)

# Run the stepwise regression 
stepGam = step.Gam(baseGam,
                   scope =
                     list("gender" = ~1 + as.factor(gender),
                          "age" = ~1 + age + s(age,3),
                          "hra" = ~1 + hra + s(hra,3),
                          "sys" = ~1 + sys + s(sys,3)),
                   family = binomial,data = icu)
```

9. What is the indicated model?

> The indicated model (Step:3) includes age and systolic blood pressure as a spline [s(sys, 3)].

10. Detach the gam package and rerun using mgcv::gam with the indicated stepwise model: use
k = 10 and bs=”cr”. What terms are statistically significant at the 0.05 level?

```{r}
# Detach the gam package and load mgcv
detach("package:gam", unload=TRUE)
library(mgcv)

# Run stepwise using mgcv
mgcvStep = gam(Y ~ age 
               + s(sys,k = 10,bs="cr"), 
               family = binomial, 
               data=icu)
summary(mgcvStep)
```
> Both age and systolic blood pressure are significant at the 0.05 level.
