---
title: "Homework 7"
author: "Skylar Liu"
date: "2024-02-24"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_template$set(figure1 = list(fig.height = 3.7, fig.width = 5),
                         figure2 = list(fig.height = 3.7, fig.width = 8))
```

Homework #7, Stat 660, Spring 2024, Due Saturday, February 25

1. Fit an ordinary logistic regression with response = CHD, and predictor = Age. Show the 
summary table. 

```{r log regres}
# clear workspace
rm(list = ls())
# set the seed
set.seed(382957)
library(HRW)
library(mgcv)

# import data
init = read.csv("~/660 - Flexible Regression/Homework/Homework6/Framingham.csv")
SBP      = rowMeans(init[, c(3:6)])
LSBP     = log(SBP - 50)
cholest  = rowMeans(init[, c(8:9)])
Lcholest = log(cholest)
data     = cbind(init[, c("CHD", "Age", "Smoker")], LSBP, Lcholest)

# fit an ordinary logistic regression
data$CHD <- as.numeric(as.character(data$CHD))

framlog <- mgcv::gam(CHD ~ Age, family = binomial(link="logit"), data = data)
summary(framlog)
```
2. In Question 1, display the fit without the data points. Just plot the model object. We know 
that when plotting the model object, mgcv::gam ignored the intercept. 

```{r plot log regres}
#Model predictor
gampred <- predict(framlog, type = "response")

# Plot the log regression model object
plot(data$Age, data$CHD, type="n",xlim = c(30,70), ylim = c(0, 1),xlab="Age", ylab="CHD", main = "CHD vs Age")
lines(gampred, lwd=2, col = "red")
```
3. In Question 1, is the fit statistically significant?

> The deviance explained for the ordinary logistic regression model is small (3.81%), suggesting that the model is a good fit. The p-values for Age is also statistically significant, with 1.12x10^-8 much smaller than 0.05.

4. In Question 1, test whether the fit is linear or quadratic versus the need to do a 
semiparametric fit, i.e., a spline in age

```{r anova}
# Fit a quadratic fit
framquad = mgcv::gam(CHD ~ poly(Age,degree=2), family = binomial(link="logit"),    
                      data = data)

# Fit a semiparametric fit
framparam = mgcv::gam(CHD ~ s(Age,bs="cr"), family = binomial(link="logit"),    
                      data = data)

# Test linear vs semiparametric
anova(framlog,framparam,test="Chisq")

# Test quadratic vs semiparametric
anova(framquad,framparam,test="Chisq")
```
>A spline is only necessary in the linear fit, with a statistically significant p-value of 0.033 which is less than 0.05. The p-value for the quadratic fit is not statistically significant at 0.2153, suggesting we don't need the spline for this case.

5. Fit a logistic gam with all the predictors but only LSBP modeled as a spline. 

```{r all predicts}
# Fit a logistic game with all predictors, LSBP as spline
framall <- mgcv::gam(CHD ~ Age + Smoker + Lcholest + s(LSBP,bs="cr"), 
                     family = binomial(link="logit"), 
                     data = data)
summary(framall)
```

>a. Quote the p-values for all of the predictors. 
P-values:
Age: 1.87e-06 
Smoker: 0.016
Lcholest: 3.61e-06
LSBP: 0.000537

>b. Answer whether the fit suggests that LSBP should be modeled as a spline.
Remember, you need to do an ANOVA for this will the null model having 
everything modeled as ordinary logistic regression. 

```{r LSBP spline test}
# Fit a logistic game with all predictors
framall2 <- mgcv::gam(CHD ~ Age + Smoker + Lcholest + LSBP, 
                     family = binomial(link="logit"), 
                     data = data)

# Test LSBP with or without spline
anova(framall2,framall,test="Chisq")
```
> LSBP does not need to be modeled as a spline; when tested against everything modeled as ordinary logistic regression, the p-value is 0.1691 which is not statistically significant at the 0.05 level.

6. Fit a logistic gam with LSBP, Lcholest and age modeled as splines. Quote the p-values for 
all predictors. Tell me which of the spline terms seem like they are worth modeling as a 
spline. Remember, you need to do an ANOVA for this will the null model having 
everything modeled as ordinary linear logistic regression, but use mgcv::gam.

```{r splines}
# Fit a logistic gam with LSBP, Lcholest, and age as splines
framsplines <- mgcv::gam(CHD ~ s(LSBP,bs="cr") + s(Lcholest,bs="cr") + s(Age,bs="cr") + Smoker, 
                     family = binomial(link="logit"), 
                     data = data)
summary(framsplines)

# Compare to ordinary linear logistic regression
anova(framall2,framsplines,test="Chisq")

# Test each spline to see which is significant
framLSBP <- mgcv::gam(CHD ~ s(LSBP,bs="cr") + Lcholest + Age + Smoker, 
                     family = binomial(link="logit"), 
                     data = data)
framLcholest <- mgcv::gam(CHD ~ LSBP + s(Lcholest,bs="cr") + Age + Smoker, 
                     family = binomial(link="logit"), 
                     data = data)
framAge <- mgcv::gam(CHD ~ LSBP + Lcholest + s(Age,bs="cr") + Smoker, 
                     family = binomial(link="logit"), 
                     data = data)

# Compare to ordinary linear logistic regression
anova(framall2,framLSBP,test="Chisq")
anova(framall2,framLcholest,test="Chisq")
anova(framall2,framAge,test="Chisq")
```
> P-values:
Smoker: 0.0181
LSBP: 0.000353
Lcholest: 6.60e-06
Age: 4.45e-05

>Lcholest is the only statistically significant spline term, with a p-value of 0.018. LSBP has a p-value of 0.169 and Age has a p-value of 0.056, which are not statistically significant at the 0.05 level.

7. This is an open-ended question with no absolutely correct answer. It will not be graded. At 
some point in life, you are going to have to summarize your results to people who do not 
care for the details of the analysis. So, having done an exhaustive and exhausting analysis, 
write a paragraph free of technical jargon about what things you think might be important 
in predicting who is at higher risk of getting coronary heart disease.

> There are many possible variables that may contribute to a hightened risk of getting coronoary heart disease. Based on our recent analysis, I have analyzed Age, Cholesterol, Smoking status, and Systolic blood pressure. From the data provided, I have observed cholesterol and age to be the greatest predictors of getting coronary heart disease. 
