---
title: "Homework 3"
author: "Skylar Liu"
date: "2024-01-27"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_template$set(figure1 = list(fig.height = 3.7, fig.width = 5),
                         figure2 = list(fig.height = 3.7, fig.width = 8))
```

Homework #3, Stat 660, Spring 2024, Due Class #4, January 29 

1. Please find the fossil data on Canvas, and download it and its documentation 

2. Do a scatterplot of the data, with X = age. As seen near the end of Lecture 1, it is distinctly 
not linear, and hence a perfect example for semiparametric regression.

```{r scatterplot, opts.label = "figure1", fig.align="center"}
# clear workspace
rm(list = ls())
# set the seed
set.seed(382957)

# import data
fossil = read.csv("~/660 - Flexible Regression/Homework/Homework3/fossil.csv")
X = fossil$Age
Y = fossil$Strontium.Ratio

# scatterplot X vs Y
plot(X, Y, type="p",pch='*',main="Fossil Age vs. Strontium Ratio",
     xlab="Age",ylab="Strontium Ratio",cex=2)
```

3. Fit a linear model and add the fit to the scatterplot

```{r linear model, opts.label = "figure1", fig.align="center"}
# linear regression of Y on X
lin = lm(Y ~ X)

# scatterplot X vs Y
plot(X, Y, type="p",pch='*',main="Fossil Age vs. Strontium Ratio",
     xlab="Age",ylab="Strontium Ratio",cex=2)
abline(lin, col="red")
```

4. Fit a quadratic model and add the fit to the scatterplot of the raw data. Does this help? Why 
or why not? 

```{r quadratic model, opts.label = "figure1", fig.align="center"}
#create a new variable for age^2
X2 = X^2

#fit quadratic regression model
quad = lm(Y ~ X + X2)

#create list of predicted happines levels using quadratic model
strontPredict = predict(quad)
ord = sort(X, index.return = T)$ix

#scatterplot of X and Y
plot(X, Y, type="p",pch='*',main="Fossil Age vs. Strontium Ratio",
     xlab="Age",ylab="Strontium Ratio",cex=2)

#add predicted lines based on quadratic regression model
lines(X[ord], strontPredict[ord], col='red')
```

> A quadratic fit does not help since a quadratic line does not fit the curved path of the scatterplot. A quadratic fit is bimodal (one knot) while this graph appears to be multi-modal (two knots).

5. Fit a cubic model and add the fit to the scatterplot of the raw data. Does this help? Why or 
why not? 

```{r cubic model, opts.label = "figure1", fig.align="center"}
#create a new variable for age^3
X3 = X^3

#fit quadratic regression model
cub = lm(Y ~ X + X2 + X3)

#create list of predicted happines levels using quadratic model
strontPredict = predict(cub)

#scatterplot of X and Y
plot(X, Y, type="p",pch='*',main="Fossil Age vs. Strontium Ratio",
     xlab="Age",ylab="Strontium Ratio",cex=2)

#add predicted lines based on quadratic regression model
lines(X[ord], strontPredict[ord], col='red')
```
> This fit is better because it follows the curves of the scatterplot. The curve follows the two knots suggested by the scatterplot.

6. Fit a quartic (x+x^2+x^3+x^4) model and add the fit to the scatterplot of the raw data. Does 
this help? Why or why not? 

```{r quartic model, opts.label = "figure1", fig.align="center"}
#create a new variable for age^4
X4 = X^4

#fit quadratic regression model
quart = lm(Y ~ X + X2 + X3 + X4)

#create list of predicted happines levels using quadratic model
strontPredict = predict(quart)

#scatterplot of X and Y
plot(X, Y, type="p",pch='*',main="Fossil Age vs. Strontium Ratio",
     xlab="Age",ylab="Strontium Ratio",cex=2)

#add predicted lines based on quadratic regression model
lines(X[ord], strontPredict[ord], col='red')
```
> This model does not really help as it's very similar to the cubic regression model. Since the cubic model follows the curves with the right amount of knots, I would prefer that model.

